Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-07-19 analysis/paper/paper.qmd
work on the background section

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. According to specific spatial and temporal contexts, the forms of European colonization and power relations varied across regions [@Jorden2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources for commercial profit. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, or entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009].

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan. Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control. However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export, involving many Indigenous communities and Han Chinese. Despite some discussions on the policies and operations of the deerskin trade, direct archaeological investigations into the experiences of local Indigenous people to these commercial activities are rare. The European impacts on local ecological conditions have been recorded in many cases in colonial settings [@Jorden2009]. Prior to the deerskin trade, deer has been a substantial part of local subsistence. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), a largest settlement in northeastern Taiwan [@citation]. Kiwulan experienced periods of European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. I will primarily focus on subsistence practices, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment. My research questions are: What were the subsistence practices of Indigenous people before and after the European presence? Can we identify shifts in these strategies in response to colonial economic demands? To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues for investigating culinary practices. Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# Colonial encounter in Northeastern Taiwan
<!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to avoid the interference in the Asian trade by the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company), which established a colony in southern Taiwan at the same time. In 1642, the Dutch defeated the Spanish and took over their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in 1662 [@Andrade2007]. The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and the collection resources they needed among different Indigenous groups. For example, the Basay people, active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@citation]. The European domination was mostly established through the conquest over indigenous villages, and the VOC further developed tribute system to request deerskins or products from indigenous people []. However, for most indigenous villages distant from the European colonial base, European colonial control was relatively less, indirect, and even unsuccessful.

<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->
Kavalan, or Yilan in the present day, in northeastern Taiwan is an example of slight European presence for most of the time. Kavalan is about 70 kilometers away from the European northern base with mountains in between. Local indigenous groups were Kavalan people, who first experienced European powers in 1632, when they were attacked by the Spanish. Later in 1647, a group of Dutch soldiers attacked many Kavalan villages and asked those Kavalan villages to pay tributes [@citation]. The Dutch also built a trading outpost in Trobiawan located near the coastal area at Yilan, but there might be not much officials at any given time. According to the Dutch census, there were more than 40 Kavalan villages with approximately 10,000 people in the 17th century in this region [@Nakamura1938], which were also mentioned by the Chinese and Japanese in later periods, demonstrating a continuous Indigenous habitation until the 19th century. It appears that the Kavalan area was in the situation where trading activities were frequent and active stimulated by the European presence in northern Taiwan, with occasional pressures from colonial powers. In general, the European control over the Kavalan must have been largely indirect with limited colonialism [@].

Before the arrival of the Europeans in northern Taiwan,

@@

Northeastern Taiwan is an especially interesting example of European and Indigenous interaction because the impact of European colonialism appears to have been much less pronounced compared to other places in this region. I define northeastern Taiwan as an area at the periphery of colonial center where the European power was indirect with a few sporadic, mostly non-violent direct contacts. This type of context has also been called pericolonialism (Acabado, 2017; Kang, 2016, pp. 31-78; Trabert, 2017). My PhD study at Kiwulan (14-19th A.D.), the biggest Indigenous settlement during the 17th century in Yilan, indicates that European contact and foreign goods resulted in an increased social differentiation in the local society through analyses of trade items, pottery production, and burial goods (Wang and Marwick, 2020a, 2020b, 2021). The results demonstrate that there was an indirect effect of European colonization in this region that could lead to changes in the organizational mode of a society that have not been identified before.
In addition to social changes, how did the interactions with Europeans and Chinese lead to changes in the local ecological system need further examination. For example, during the Dutch colonial period, deer hide was especially demanded in the Dutch colonization period as payment for tributes due to its high value for export to Japan, which led to intensified deer hunting in historical Taiwan (Andrade, 2007; Koo, 2011). While the commercial activities and a substantial decrease in Taiwan’s deer population were well documented, much less is known about how this historical consequence can be situated in a local ecological framework. Did the participation in these trade activities result in a change in Indigenous subsistence behaviors or the human-environment interactions? What was the response of local Indigenous people to the decrease of deer population? Those questions are important because they can help us to trace back the Indigenous cultures and their resilience under colonial interference since the protohistoric Taiwan.

>>>>

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jorden2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources for commercial profit. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009; @Alexander1998].

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan. Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control. However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export in a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001]. Despite some discussions on the policies and operations of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people to the commercial activities are rare. The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jorden2009]. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), a largest settlement in northeastern Taiwan [@Chen2007]. Kiwulan experienced periods of European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. I will primarily focus on subsistence practices, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment. My research questions are: What were the subsistence practices of Indigenous people before and after the European presence? Can we identify shifts in these strategies in response to colonial economic demands? To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues for investigating culinary practices. Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# Colonial encounter in Northeastern Taiwan
<!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in Asian market, which established a colony in southern Taiwan two years before the Spanish in the north. In 1642, the Dutch defeated the Spanish and occupied their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in Taiwan in 1662 [@Andrade2007]. The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and resources collections among different Indigenous groups. For example, the Basay people, an indigenous group active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@Kang2016]. The colonial domination was mostly established through the conquest over indigenous villages, but the colonial control varied depending on the location of indigenous settlements. For most indigenous villages distant from the European colonial base, the colonial control was relatively less, indirect, and even unsuccessful.
<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->

Kavalan, or the present day Yilan, in northeastern Taiwan is an example of slight European presence for most of the time. Kavalan is about 70 kilometers away from the European northern fortress with mountains in between. Local indigenous groups were Kavalan people. According to the Dutch census, there were more than 40 Kavalan villages with approximately 10,000 people in the 17th century [@Nakamura1938], which were also mentioned by the Chinese and Japanese in later periods, suggesting a continuous indigenous habitation until the 19th century. Kavalan people first experienced European powers in 1632, when they were attacked by Spanish traders [@citation]. Later in 1647, a group of Dutch soldiers attacked several Kavalan villages and asked those defeated villages to pay tributes [@Kang2016]. According to Dutch written record, the VOC had built a trading outpost in Trobiawan, located near the coastal area in Yilan, but there might be not much officials at any given time. It appears that the Kavalan area was in the situation where trading activities were frequent and active stimulated by the European presence in northern Taiwan, with occasional pressures from colonial powers. The European control over the Kavalan must have been largely indirect and limited, which allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes requested by the VOC, such as deerskins or products [@Kang2016].

Before the arrival of the Europeans in northern Taiwan, there had been a regional exchange network involving Chinese merchants, Japanese traders, and indigenous groups in northern and northeastern Taiwan since the 14th century [@citation]. The local indigenous groups traded

@@

Northeastern Taiwan is an especially interesting example of European and Indigenous interaction because the impact of European colonialism appears to have been much less pronounced compared to other places in this region. I define northeastern Taiwan as an area at the periphery of colonial center where the European power was indirect with a few sporadic, mostly non-violent direct contacts. This type of context has also been called pericolonialism (Acabado, 2017; Kang, 2016, pp. 31-78; Trabert, 2017). My PhD study at Kiwulan (14-19th A.D.), the biggest Indigenous settlement during the 17th century in Yilan, indicates that European contact and foreign goods resulted in an increased social differentiation in the local society through analyses of trade items, pottery production, and burial goods (Wang and Marwick, 2020a, 2020b, 2021). The results demonstrate that there was an indirect effect of European colonization in this region that could lead to changes in the organizational mode of a society that have not been identified before.

In addition to social changes, how did the interactions with Europeans and Chinese lead to changes in the local ecological system need further examination. For example, during the Dutch colonial period, deer hide was especially demanded in the Dutch colonization period as payment for tributes due to its high value for export to Japan, which led to intensified deer hunting in historical Taiwan (Andrade, 2007; Koo, 2011). While the commercial activities and a substantial decrease in Taiwan’s deer population were well documented, much less is known about how this historical consequence can be situated in a local ecological framework. Did the participation in these trade activities result in a change in Indigenous subsistence behaviors or the human-environment interactions? What was the response of local Indigenous people to the decrease of deer population? Those questions are important because they can help us to trace back the Indigenous cultures and their resilience under colonial interference since the protohistoric Taiwan.
